---
title: "QSIDE 2022"
author: "Anna Tsai"
date: "`r Sys.Date()`"
output: html_document
---
## Import Libraries
```{r}
library(tidyverse)
library(ggplot2)
```

## Get Data
```{r}
offender_legal <- read.csv('/Users/actt/Documents/QSIDE Datathon 2022/Offender Legal.csv')
offender_demographics <- read.csv('/Users/actt/Documents/QSIDE Datathon 2022/Offender Demographics.csv')
```

## Join Data
```{r}
offender_merged <- offender_legal %>% 
  inner_join(offender_demographics, by = 'DOC_ID')
```

## Exploratory Analysis
```{r}
ggplot(data = offender_merged, mapping = (aes(x= TOTAL_SENTENCED_MONTHS, y = ..density..))) +
  geom_histogram(color = 1, fill = 'white', na.rm = TRUE) +
  facet_wrap(~ RACE, scale = 'free_y')
```

It seems unusual that there would be so many sentences of zero months. One possible explanation is that from inspecting the data we can see that sometimes life sentences are coded as 0 months and other times they are coded as several thousand months. 


```{r}
sentence_summary <- offender_merged %>% 
  filter(RACE == 'Native American' & !is.na(TOTAL_SENTENCED_MONTHS)) %>%
  group_by(COUNTY, OFFENSE, .add= TRUE)  %>%
  summarize(mean = mean(TOTAL_SENTENCED_MONTHS), count = n())
```


## Time Series for Native Americans
```{r}
#Convert PRONOUNCEMENT_DATE from string to date and extract year
offender_merged <- offender_merged %>%
  mutate(YEAR=as.Date(PRONOUNCEMENT_DATE, format = "%m/%d/%Y"))

offender_merged <- offender_merged %>%
  mutate(YEAR=format(YEAR, format = "%Y"))

#Filter by race and group by year
filtered_offender_by_year <- offender_merged %>%
  filter(RACE == 'Native American' &!is.na(YEAR)) %>%
  group_by(YEAR) %>%
  summarize(TOTAL_SENTENCES = n())
```

```{r}
#Plot data
ggplot(data = filtered_offender_by_year, mapping = (aes(x = YEAR, y = TOTAL_SENTENCES)), ) +
  geom_point() +
  geom_vline(xintercept = '2013', color = 'red') +
  geom_vline(xintercept = '2010', color = 'red') +
  theme(axis.text.x = element_text(margin = unit(c(0, 0, 0, 0), "mm"), angle = 90)) +
  xlab('Year')+
  ylab('Total Sentences')+
  labs(title = 'Total Sentences By Year for Native Americans in Montana', caption = 'The red lines represent years when siginficant legislation affecting tribal sovereignty was passed')

ggsave('Total Sentences by Year.png')
```


We can see an increase in slope around 2000/2001 and some increase in slope around the time that significant legislation was passed.

## Time Series for All Races
```{r}
#Convert PRONOUNCEMENT_DATE from string to date and extract year
offender_merged <- offender_merged %>%
  mutate(YEAR=as.Date(PRONOUNCEMENT_DATE, format = "%m/%d/%Y"))

offender_merged <- offender_merged %>%
  mutate(YEAR=format(YEAR, format = "%Y"))

#Group by year
sentences_by_year <- offender_merged %>%
  filter(!is.na(YEAR)) %>%
  group_by(RACE, YEAR, .add= TRUE) %>%
  summarize(TOTAL_SENTENCES = n())
```

```{r}
#Plot data
ggplot(data = filtered_offender_by_year, mapping = (aes(x = YEAR, y = TOTAL_SENTENCES, color = RACE)), ) +
  geom_point() +
  theme(axis.text.x = element_text(margin = unit(c(0, 0, 0, 0), "mm"), angle = 90)) +
  xlab('Year')+
  ylab('Total Sentences')+
  labs(title = 'Total Sentences By Year in Montana')

ggsave('Total Sentences by Year_All Races.png')
```

We can see here that we do not see similar increases for Asian or Pacific Islander and Black populations but do see a similar increase in slope for Native American and White populations.